
# Social Network Analysis tutorial
The 2nd IPSA-HSE Summer School for Methods of Political & Social Research. 
St. Petersburg, Russia.

## Read libraries

```{r}
library(igraph)
library(tidyverse)
library(RColorBrewer)
library(visNetwork)
library(ggrepel)
```

## Read data

In this tutorial we will use two datasets constructed from the results of [European Parliament Election Study 2009](https://dbk.gesis.org/dbksearch/SDesc2.asp?no=5055&ll=10&af=&nf=1&db=e&search=&search2=&notabs=1). This study was about general political attitudes and behaviour and media usage of EU citizens. Our datasets contains only respondents from Great Britan.

graph_gb_papers.csv - is a weigthed edgelist. Each row in this file is an edge of a network, where nodes are newpapers. Edge between two nodes exists in case if one person reads both of them. Third column in this table is a weight of edge - number of common readers.

```{r}
net_edgelist = read.csv("graph_gb_papers.csv")
View(net_edgelist)
```

gb_data_survey.csv - is table with additional information about the readers. Each row is a pair "respondent - newspaper". In case if person reads two papers, there will be 2 rows with his characteristics, by one for each newspapers. 

Variables:
id - identification number of respondent
region - region of living
problem1 - public issue refered as the most important for Great Britan
party - party affiliation
days_a_week_news - how many days a week respondent follow political news
left_right - political orientation (0 - left, 10 - right)
membership_eu - opinion about Britan's membership in European union
sex - ...
birth_year - ...
social_class - ...
urban - level of living area urbanization
religion - religion affilation
wealth - level of wealth (1 - extremly poor, 7 - extremly rich)
newspaper - newspaper the respondent used to read


```{r}
meta = read.csv("gb_data_survey.csv")
View(meta)
```

## Load network data to igraph

```{r}
net = graph_from_data_frame(net_edgelist, directed = FALSE)
net
```

## Static visualization - first try

Looks ugly!

```{r}
plot(net)
```

Lets remove titles and reduce the node size. 


```{r}
# задаем параметры
plot(net, 
     vertex.size = 4, 
     vertex.label = NA)
# что можно менять
# узлы - связи - подписи
# размер - цвет - форма
# + расположение
```

```{r}
my_layout = layout.fruchterman.reingold(net)

plot(net, 
     vertex.size = 4, 
     vertex.color = "steelblue1",
     layout = my_layout, 
     vertex.label = NA,
     edge.curved = 0.2)
```

```{r}
E(net)$width = log(E(net)$weight+0.5)*2

plot(net, 
     vertex.size = 4, 
     vertex.color = "steelblue1",
     layout = my_layout, 
     vertex.label = NA,
     edge.curved = 0.2,
     edge.width = E(net)$width)
```


# centrality measures

```{r}
V(net)$degree = degree(net)
V(net)$size = log(V(net)$degree)*2

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = "steelblue1",
     layout = my_layout, 
     vertex.label = NA,
     edge.curved = 0.2)
```

# убрать редкие газеты

```{r}

rare_papers = meta %>% 
  group_by(newspaper) %>% 
  summarize(count = n()) %>% 
  filter(count < 3)

rare_papers$newspaper = as.character(rare_papers$newspaper)

net2 = delete.vertices(net, V(net)$name %in% rare_papers$newspaper)
my_layout2 = layout.fruchterman.reingold(net2)

plot(net2,
     vertex.size = V(net2)$size, 
     vertex.color = "steelblue1",
     layout = my_layout2, 
     vertex.label = NA,
     edge.curved = 0.2,
     edge.width = E(net2)$width)
```

```{r}
quantile(V(net)$degree, 0.9)
V(net)$label = ifelse(V(net)$degree > quantile(V(net)$degree, 0.9), V(net)$name, NA)

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = "steelblue1",
     layout = my_layout,
     edge.curved = 0.2, 
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```

# community detection

# interpretation
```{r}
meta$left_right[meta$left_right %in% c(77, 88)] = NA
meta$wealth[meta$wealth %in% c(77, 88)] = NA

attr_net = meta %>% 
  group_by(newspaper) %>% 
  summarize(
    left_right_median = round(median(wealth, na.rm = T)),
    count = n())
  
attr_net = meta %>% 
  group_by(newspaper, social_class) %>% 
  summarize(count = n()) %>% 
  top_n(n = 1, wt = count) %>% 
  select(-count) %>% 
  right_join(attr_net)

attr_net = attr_net[!duplicated(attr_net$newspaper),]

colors = RColorBrewer::brewer.pal(name = "Set1", n = 9)
attr_net$color = colors[as.numeric(attr_net$social_class)]

attr_net = attr_net %>% 
  filter(newspaper %in% V(net)$name)

attr_net$newspaper = as.character(attr_net$newspaper)
attr_net$social_class = as.character(attr_net$social_class)                

V(net)$color = attr_net$color[match(attr_net$newspaper, V(net)$name)]
V(net)$social_class = attr_net$social_class[match(attr_net$newspaper, V(net)$name)]

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = V(net)$color,
     layout = my_layout,
     edge.curved = 0.2, 
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```

# community
```{r}
clust_net = cluster_edge_betweenness(net)
V(net)$color = colors[clust_net$membership]

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = V(net)$color,
     layout = my_layout,
     edge.curved = 0.2, 
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```


```{r}
# https://datastorm-open.github.io/visNetwork/
data <- toVisNetworkData(net)

visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout()
```

```{r}
data$nodes$size = (data$nodes$size+0.75) * 5
data$edges$width = data$edges$width * 3
data$nodes$title = data$nodes$label

visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout() %>% 
  visEdges(color = list("opacity" = 0.25))

```

```{r}


visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout() %>% 
  visEdges(color = list("opacity" = 0.25)) %>% 
  visOptions(highlightNearest = TRUE,
             selectedBy = list(variable = "social_class", multiple = T)) 
  

```

```{r}

meta_clust = data.frame(
  newspaper = V(net)$name,
  cluster = V(net)$color,
  stringsAsFactors = FALSE
)

meta = left_join(meta, meta_clust)
```

```{r}
meta_aggr = meta %>%  
  group_by(newspaper, cluster) %>% 
  summarize(wealth = mean(wealth, na.rm = T), 
            left_right = mean(left_right, na.rm = T),
            count_readers = n()) %>% 
  filter(count_readers > 2) %>% 
  filter(!is.na(cluster))

meta_aggr = meta_aggr %>% 
  group_by(cluster) %>% 
  summarize(cluster_count = n()) %>% 
  right_join(meta_aggr) %>% 
  filter(cluster_count > 1)
  
ggplot() +
  geom_point(data = meta_aggr, 
              aes(x = wealth, y = left_right, 
                  size = log(count_readers), 
                  fill = cluster), 
              color = "black", 
              alpha = 0.35, shape = 21, stroke = 1) +
  scale_size(range = c(4, 10)) +
  xlab("Wealth") +
  ylab("Left - Right") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#4DAF4A" = "#4DAF4A", 
                               "#E41A1C" = "#E41A1C",
                               "#377EB8" = "#377EB8")) +
  scale_color_manual(values = c("#4DAF4A" = "#4DAF4A", 
                               "#E41A1C" = "#E41A1C",
                               "#377EB8" = "#377EB8")) +
  geom_text_repel(data = top_n(meta_aggr, 15, count_readers), 
                  aes(x = wealth, y = left_right, 
                  color = cluster, label = newspaper),
                  size = 5, point.padding = 0.1)
  

```
