
# Social Network Analysis tutorial
The 2nd IPSA-HSE Summer School for Methods of Political & Social Research. 
St. Petersburg, Russia.

## Read libraries

```{r}
library(igraph)
library(tidyverse)
library(RColorBrewer)
library(visNetwork)
library(ggrepel)

# set plotting area options
par(mar = c(0,0,0,0))
```

## Read data

In this tutorial we will use two datasets constructed from the results of [European Parliament Election Study 2009](https://dbk.gesis.org/dbksearch/SDesc2.asp?no=5055&ll=10&af=&nf=1&db=e&search=&search2=&notabs=1). This study was about general political attitudes and behaviour and media usage of EU citizens. Our datasets contains only respondents from Great Britan.

graph_gb_papers.csv - is a weigthed edgelist. Each row in this file is an edge of a network, where nodes are newpapers. Edge between two nodes exists in case if one person reads both of them. Third column in this table is a weight of edge - number of common readers.

```{r}
net_edgelist = read.csv("graph_gb_papers.csv")
View(net_edgelist)
```

gb_data_survey.csv - is table with additional information about the readers. Each row is a pair "respondent - newspaper". In case if person reads two papers, there will be 2 rows with his characteristics, by one for each newspapers. 

Variables:  
id - identification number of respondent  
region - region of living  
problem1 - public issue refered as the most important for Great Britan  
party - party affiliation  
days_a_week_news - how many days a week respondent follow political news  
left_right - political orientation (0 - left, 10 - right)  
membership_eu - opinion about Britan's membership in European union  
sex - ...  
birth_year - ...  
social_class - ...  
urban - level of living area urbanization  
religion - religion affilation  
wealth - level of wealth (1 - extremly poor, 7 - extremly rich)  
newspaper - newspaper the respondent used to read  

```{r}
meta = read.csv("gb_data_survey.csv", stringsAsFactors = FALSE)
View(meta)
```

## Load network data to igraph

```{r}
net = graph_from_data_frame(net_edgelist, directed = FALSE)
net
```

## Static visualization - first try

Looks ugly!

```{r}
plot(net)
```

Lets remove titles and reduce the node size. 

In igraph we can edit appearance of nodes with parameters starting from *vertex.*, while edges' visualization might be assesed by with prefix *edge.*.

```{r}
plot(net, 
     vertex.size = 4, 
     vertex.color = "steelblue1", 
     vertex.label = NA,
     edge.curved = 0.2)
```

Main elements of network visualization are nodes, edges, labels and layout. Main appearance paramentes: color, size, shape. 

We can recompute the layout of network.

```{r}
my_layout = layout.fruchterman.reingold(net)
# you may try another layout e.g. layout.kamada.kawai()

plot(net, 
     vertex.size = 4, 
     vertex.color = "steelblue1",
     vertex.label = NA,
     edge.curved = 0.2,
     layout = my_layout)
```

We can use the width of edges to show how many common reader the newspapers have. Whan we loaded our edgelist to igraph, the weigths were automatically detected. In igraph we can save attributes of the nodes and edges. They can be accesed with the following structure.

V(*graph_object_name*)$*attribute_name* (V - stands for vertices)  
E(*graph_object_name*)$*attribute_name* (E - stands for edges)

Lets transform the weights to have less ranged distribution. 

```{r}
E(net)$width = log(E(net)$weight+0.5)*2

plot(net, 
     vertex.size = 4, 
     vertex.color = "steelblue1", 
     vertex.label = NA,
     edge.curved = 0.2,
     layout = my_layout,
     edge.width = E(net)$width)
```

One of the practices to improve readability of the graph is to remove "non-important" nodes or edges.

Lets compute how many readers have each newspaper in our sample and remove those of them whch have less than 3.

```{r}
rare_papers = meta %>% 
  group_by(newspaper) %>% 
  summarize(count = n()) %>% 
  filter(count < 3)

net2 = delete.vertices(net, V(net)$name %in% rare_papers$newspaper)

# recompute layout
my_layout2 = layout.fruchterman.reingold(net2)

plot(net2,
     vertex.size = V(net2)$size, 
     vertex.color = "steelblue1",
     vertex.label = NA,
     edge.curved = 0.2,
     layout = my_layout2,
     edge.width = E(net2)$width)
```

## Centrality measures

There are different centrality measures for nodes in network analysis.   
degree - number of adjacent edges   
closeness - how many steps is required to access every other node from a given node  
betweenness - (roughly) the number of shortest paths going through a node  

```{r}
V(net)$degree = degree(net)
V(net)$size = log(V(net)$degree)*2

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = "steelblue1", 
     vertex.label = NA,
     edge.curved = 0.2,
     layout = my_layout,
     edge.width = E(net)$width)
```

We could add titles for a particular set of nodes. For example, for those which are in top-10% list of nodes by degree.

```{r}
quantile(V(net)$degree, 0.9)
V(net)$label = ifelse(V(net)$degree > quantile(V(net)$degree, 0.9), V(net)$name, NA)

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = "steelblue1",
     edge.curved = 0.2, 
     layout = my_layout,
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```

We could re-color the nodes to show some additional dimension in our data. Which journals are mostly read by conservatives? 

```{r}
attr_net = meta %>% 
  group_by(newspaper, party) %>% 
  summarize(count = n()) %>% 
  top_n(n = 1, wt = count) %>% 
  select(-count)

# select one party if there are two with the same number of readers
attr_net = attr_net[!duplicated(attr_net$newspaper),]

# select qualitive color sclae (http://colorbrewer2.org)
colors = RColorBrewer::brewer.pal(name = "Set1", n = 9)

attr_net$party_num = attr_net$party %>% 
  as.factor() %>% 
  as.numeric()

attr_net$color = colors[attr_net$party_num]

# filter out newspapers which are not in the network
attr_net = attr_net %>% 
  filter(newspaper %in% V(net)$name)

# add new color to graph's nodes assigned by the dominant party among readers
V(net)$color = attr_net$color[match(V(net)$name, attr_net$newspaper)]

# plot network
plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = V(net)$color,
     layout = my_layout,
     edge.curved = 0.2, 
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```

## Community detection

Many networks consist of modules which are densely connected themselves but sparsely connected to other modules. 

```{r}
clust_net = cluster_edge_betweenness(net)
# you may try another community detection algorithms e.g. cluster_louvain()

V(net)$color = colors[clust_net$membership]

plot(net, 
     vertex.size = V(net)$size, 
     vertex.color = V(net)$color,
     layout = my_layout,
     edge.curved = 0.2, 
     edge.width = E(net)$width,
     vertex.label.cex = 0.75, 
     vertex.label = V(net)$label, 
     vertex.label.color = "black")
```

To interpret how these groups of newspapers were formed we can refer to external data about their readers.

Let check how these clusters looks in two-dimensional space on political orientation of readers and their wealth.

```{r}
# extract newspaper title and community color 
meta_clust = data.frame(
  newspaper = V(net)$name,
  cluster = V(net)$color,
  stringsAsFactors = FALSE
)

# add community color to metadata
meta = left_join(meta, meta_clust)

# missings were coded as 77 and 88 
meta$left_right[meta$left_right %in% c(77, 88)] = NA
meta$wealth[meta$wealth %in% c(77, 88)] = NA

# compute mean political orientation and welath for each newspaper
meta_aggr = meta %>%  
  group_by(newspaper, cluster) %>% 
  summarize(wealth = mean(wealth, na.rm = T), 
            left_right = mean(left_right, na.rm = T),
            count_readers = n())

# remove newspapaers with less than 3 readers 
# and those which are not in the network and not assigned to any community 
meta_aggr = meta_aggr %>% 
  filter(count_readers > 2) %>% 
  filter(!is.na(cluster))

# remove communities consists of less than 2 newspapers
meta_aggr = meta_aggr %>% 
  group_by(cluster) %>% 
  summarize(cluster_count = n()) %>% 
  right_join(meta_aggr) %>% 
  filter(cluster_count > 1)
  
# ggplot empty layer
p = ggplot() 

# add points on the plot
# each point is a newspapaer
# size is logarithmed number of readers 
# color is community
p = p + geom_point(data = meta_aggr, 
                   aes(x = left_right, y = wealth, 
                       size = log(count_readers), 
                       fill = cluster), 
                   color = "black", alpha = 0.35, 
                   shape = 21, stroke = 1) 

p

# change size scale
# change axis titles
# add appearence theme 
# remove legend
p = p +
  scale_size(range = c(4, 10)) +
  ylab("Wealth") +
  xlab("Left - Right") +
  theme_minimal() +
  theme(legend.position = "none")

p

# color codes were detected as just levels of factor variable
# set the same colors as on the network visualization
p = p +
  scale_fill_manual(values = c("#4DAF4A" = "#4DAF4A", 
                               "#E41A1C" = "#E41A1C",
                               "#377EB8" = "#377EB8")) +
  scale_color_manual(values = c("#4DAF4A" = "#4DAF4A", 
                               "#E41A1C" = "#E41A1C",
                               "#377EB8" = "#377EB8")) 

p

# add titles of the top 15 newspapers by number of readers
p = p +
  geom_text_repel(data = top_n(meta_aggr, 15, count_readers), 
                  aes(x = left_right, y = wealth, 
                  color = cluster, label = newspaper),
                  size = 5, point.padding = 0.1)
  
p
```

## Interactive visualization

visNetwork is an R package for network visualization. It allows to convert network from igraph without difficulties. 

```{r}
# https://datastorm-open.github.io/visNetwork/
data <- toVisNetworkData(net)

visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout()
```

Size of nodes and some other appearance parameters are not looking good. Lets ajust it a little bit. Also we can add title for nodes, which will appear as a popup when the mouse will hover over node. 

```{r}
data$nodes$size = (data$nodes$size+0.75) * 5
data$edges$width = data$edges$width * 3
data$nodes$title = data$nodes$label

visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout() %>% 
  visEdges(color = list("opacity" = 0.25))
```

Moreover, visNetwork allows to add various interactivity. For exapmle to highlight nodes with particular characteristic.

```{r}
data$nodes$party = attr_net$party[match(V(net)$name, attr_net$newspaper)]

visNetwork(nodes = data$nodes, edges = data$edges) %>% 
  visIgraphLayout() %>% 
  visEdges(color = list("opacity" = 0.25)) %>% 
  visOptions(highlightNearest = TRUE,
             selectedBy = list(variable = "party", multiple = T)) 
```


